A large-scale study of polymorphism in R

### 2019-03-15 paper/main.tex
accidentally deleted the bit about type systems, i put it back in in the results section for now

% %
% %
% %
% %
% \subsection{Front-End Data Collection}

% The following describes our data collection methodology:

% \begin{enumerate}
%     \item use the {\tt genthat} tracer to run tests, vignettes, and examples for a package, and optionally repeat for package's reverse dependencies;
%     \item gather type and attribute information of all arguments and return values from all calls to all functions by looking at generated trace data;
%     \item per function, consolidate information to generate a type signature.
% \end{enumerate}

% In order to get a global picture of R usage, we do the above over \textit{all} CRAN and Bioconductor packages, and aggregate type and attribute signature results over all of them.

%
%
%
%
\subsection{Back-End Data Collection}

\AT{Do we want to go down this route with this paper?
This would be Konrad's bit, if we want to try to stick our data together.}

@@

>>>>

@@

%
%
%
%
\subsection{Types in R}

\AT{may not be the correct place for this. i think i accidentally deleted it.}

R is a unityped language, and as such is trivially monomorphic, and this notion of monomorphism is far from useful.
We are interested in expanding the notion of ``type'' in R, and thankfully the language has some tools on offer.
\AT{this is discussed later on}

%
%
\subsubsection{What are possible type systems?}

Great question.
Some things:

\begin{itemize}

\item {\it The Obvious Type System (OTS)}: this type system corresponds directly with the type information that one can fetch using the {\tt typeof}, {\tt attributes}, and {\tt class} functions.
In this system, there is no distinction between vectors and scalars, and lists of differently-typed elements (e.g., a list of integers vs. a list of strings), and there is a distinction between integers and doubles, and the type of NULL and all other types.
This type system isn't particularly informative to a programmer, though it does reflect the treatment of values in the R runtime, for the {\tt typeof} a value is the type that the value has according to the R internals.

\item {\it The Fine-Grained Type System (FGTS)}: this type systems includes a number of relevant additions.
Chiefly, it distinguishes between vectors and scalars, as it is conceivable that boxing scalars as single-element vectors might be overkill.
In addition, list types are parameterized over the types of their elements (e.g., list<integer> vs. list<character>).
Finally, NAs ``in the wild'' are given the type {\tt raw\_NA}, as {\tt typeof(NA) = logical} in R.

\item {\it The Null-and-NA-Free Type System (NNTS)}: most languages consider NULLs to inhabit all types, and R does so for NAs \AT{elaborate}.
This gives us ... .

\end{itemize}

Of course, one could conceive of a more fine-grained type system than FGTS, but it's unclear that we would gain much in doing so.
Indeed, there is a space to explore in slightly less fine-grained type systems, which make modifications including:

\begin{itemize}

\item {\bf collapsing vectors and scalars}: this yields a type system slightly closer to OTS.

\item {\bf rolling NULL into other types}: in most languages, NULL is an inhabitant of all types, and this adjustment seems interesting.
Note that this goes beyond even OTS, as in R {\tt typeof(NULL) = NULL}.

\item {\bf rolling NA into other types}: \AT{i should try this}
similarly, NA is an inhabitant of all types, though NA standing alone has type {\tt logical}.
FGTS distinguishes such NAs with the {\tt raw\_NA} type, and if instead we consider NA to inhabit all types we may \AT{win}.

\item {\bf combining integers and doubles}: as in most languages, doubles and ints can stand-in for each other in nearly every scenario.
Even if not, it is trivial to convert from one to the other, particularly converting ints to doubles.
Aside, in R a double can be used to index a list, and R will floor it when performing the lookup.

\end{itemize}
